Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{J}} %%% there's no difference, right? --JB
%% (I find \mathcalx{I} ugly, with its two horizontal bars, hence the J)
\newcommand{\J}{\mathcalx{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcalx{W}}

@@

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, we claim that $\conv(\varphi)$ is
equisatisfiable to $\varphi$ for any input $\varphi$. The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv(
\varphi )$ implies the existence (resp.\ nonexistence) of a model for
$\varphi$. Moreover, information in models for $\conv( \varphi )$ contains
relevant information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$,
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$
as $14$. In fact, we claim that there exists a model of $\varphi$ that
coincides with each model of $\conv( \varphi )$ on its interpretation of all
functions in $\sfuns{\Sigma}$ other than those in $\sfundefs{\Sigma}$, i.e. $\{
s \}$ in the previous example. From a practical point of view, this is not an
issue, because $s$ is the very function that was explicitly defined by the
user, and hence already has an intended interpretation.

\begin{theorem}
If $\varphi$ is a $\Sigma$-formula that is in definitional form with respect to $\sfundefs{\Sigma}$,
and the set of function definitions $\mathcal{D}$ corresponding to functions in $\sfundefs{\Sigma}$ is admissible,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists a : {\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$,
since by our construction of $\J$ there exists an $a$ such that $a^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ppos )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

@@

>>>>

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{J}} %%% there's no difference, right? --JB
%% (I find \mathcalx{I} ugly, with its two horizontal bars, hence the J)
\newcommand{\J}{\mathcalx{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
%%% @ANDY: I'm trying "A" like "abstraction", again in sync with the abstract--concrete
%%% terminology from abstract interpretation
\newcommand{\conv}{\mathcalx{A}}

@@

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, %we claim that
$\conv(\varphi)$ is
equisatisfiable with $\varphi$ for any input $\varphi$. The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable or unsatisfiable response from an SMT solver on input $\conv(
\varphi )$ implies the existence or nonexistence of a model for
$\varphi$. Moreover, information in models for $\conv( \varphi )$ contains
pertinent information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) described above interprets $\con{c}$ as $14$,
and there exists a model of formula~(\ref{eq:ex-before}) that also interprets $\con{c}$
as $14$. In general, there exists a model of $\varphi$ that
coincides with each model of $\conv( \varphi )$ on its interpretation of all
functions in $\sfuns{\Sigma} - \sfundefs{\Sigma}$, i.e., $\{
\con{s} \}$ in the example.

%%% @ANDY: I don't get the following comment --jb
%
%From a practical point of view, this is not an
%issue, because $s$ is the very function that was explicitly defined by the
%user, and hence already has an intended interpretation.

\begin{theorem}
If $\varphi$ is a\/ $\Sigma$-formula that is in definitional form with respect to\/
$\sfundefs{\Sigma}$
and the set of function definitions $\mathcal{D}$ corresponding to functions in
$\sfundefs{\Sigma}$ is admissible, then $\varphi$ and $\conv( \varphi )$ are
equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists a : {\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$,
since by our construction of $\J$ there exists an $a$ such that $a^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ppos )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

@@

  * abstract interpretation
